---
title: Simulating COVID-19 interventions with R
author: Tim Churches
date: '2020-03-21'
output:
  html_document:
    self_contained: false
---


```{r installation, include=FALSE, eval=TRUE}
for (pkg in c("tidyverse", "magrittr", "lubridate", "knitr", 
              "devtools", "DiagrammeR", "EpiModel", 
              "parallel", "foreach", "tictoc", "patchwork")) 
  if (!requireNamespace(pkg)) install.packages(pkg)

```

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE,
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE)

library(tidyverse)
library(patchwork)
library(magrittr)
library(lubridate)
library(knitr)
library(tictoc)
suppressMessages(library(EpiModel))
library(DiagrammeR)
library(devtools)
library(parallel)
library(foreach)

tic()
```

# Source code for The Conversation article, March 2020

Please read the technical blog posts [here](https://timchurches.github.io/blog/posts/2020-03-10-modelling-the-effects-of-public-health-interventions-on-covid-19-transmission-part-1/) and  [here](https://timchurches.github.io/blog/posts/2020-03-18-modelling-the-effects-of-public-health-interventions-on-covid-19-transmission-part-2/) before attempting to use the code below.

## Simulation model compartment diagram

```{r, echo=FALSE, eval=TRUE, message=FALSE, layout="l-page"}
grViz("
digraph SEIQHRF {

  # a 'graph' statement
  graph [overlap = false, fontsize = 10] #, rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  S[label='S=Susceptible'];
  E[label='E=Exposed and infected,\nasymptomatic,\npotentially infectious'];
  I[label='I=Infected and infectious'];
  Q[label='Q=(Self-)isolated\n(infectious)'];
  H[label='H=Requires\nhospitalisation'];
  R[label='R=Recovered/immune'];
  F[label='F=Case fatality']

  # several 'edge' statements
  S->E
  I->S[style='dashed']
  E->I
  E->S[style='dashed']
  I->Q
  Q->S[style='dashed']
  I->R
  I->H
  H->F
  H->R
  Q->R
  Q->H
}
")
```

# Baseline simulation

## Load EpiModel extensions

```{r, echo=TRUE, eval=TRUE, message=FALSE}
source_files <- c("_icm.mod.init.seiqhrf.R",
                  "_icm.mod.status.seiqhrf.R",
                  "_icm.mod.vital.seiqhrf.R",
                  "_icm.control.seiqhrf.R",
                  "_icm.utils.seiqhrf.R",
                  "_icm.saveout.seiqhrf.R",
                  "_icm.icm.seiqhrf.R")

src_path <- paste0("./_posts/2020-03-18-modelling-the-effects-of-public-health-",
            "interventions-on-covid-19-transmission-part-2/")

gist_url <- "https://gist.github.com/timchurches/92073d0ea75cfbd387f91f7c6e624bd7"

local_source <- FALSE

for (source_file in source_files) {
  if (local_source) {
    source(paste(src_path, source_file, sep=""))
  } else {
    source_gist(gist_url, filename=source_file)
  }
}
```

## Simulation harness function with baseline defaults

```{r, echo=TRUE, eval=TRUE}
# function to set-up and run the baseline simulations

hc_scaler <- 10 # 1 for 10,000 pop, 10 for 100,000 pop, 100 for 1,000,000 pop

simulate <- function(# control.icm params
                     type = "SEIQHRF", 
                     nsteps = 366, 
                     nsims = 8,
                     ncores = 4,
                     prog.rand = FALSE,
                     rec.rand = FALSE,
                     fat.rand = TRUE,
                     quar.rand = FALSE,
                     hosp.rand = FALSE,
                     disch.rand = TRUE,
                     infection.FUN = infection.seiqhrf.icm,
                     recovery.FUN = progress.seiqhrf.icm,
                     departures.FUN = departures.seiqhrf.icm,
                     arrivals.FUN = arrivals.icm,
                     get_prev.FUN = get_prev.seiqhrf.icm,
                     # init.icm params
                     s.num = 99970, # 999970
                     e.num=0,
                     i.num = 3, # 30
                     q.num=0,
                     h.num=0,
                     r.num = 0,
                     f.num = 0,
                     # param.icm params
                     inf.prob.e = 0.02, 
                     act.rate.e = 10,
                     inf.prob.i = 0.05, 
                     act.rate.i = 10,
                     inf.prob.q = 0.02, 
                     act.rate.q = 2.5,                    
                     quar.rate = 1/30, 
                     hosp.rate = 1/100,
                     disch.rate = 1/15,
                     prog.rate = 1/10,
                     prog.dist.scale = 5,
                     prog.dist.shape = 1.5,
                     rec.rate = 1/20,
                     rec.dist.scale = 35,
                     rec.dist.shape = 1.5,
                     fat.rate.base = 1/50,
                     hosp.cap = 40*hc_scaler, # 4000 replace red ref line too
                     fat.rate.overcap = 1/25,
                     fat.tcoeff = 0.5,
                     vital = TRUE,
                     a.rate = (10.5/365)/1000, 
                     a.prop.e = 0.01,
                     a.prop.i = 0.001,
                     a.prop.q = 0.01,
                     ds.rate = (7/365)/1000, 
                     de.rate = (7/365)/1000, 
                     di.rate = (7/365)/1000,
                     dq.rate = (7/365)/1000,
                     dh.rate = (20/365)/1000,
                     dr.rate = (7/365)/1000,
                     out="mean"
                    ) {

  control <- control.icm(type = type, 
                         nsteps = nsteps, 
                         nsims = nsims,
                         ncores = ncores,
                         prog.rand = prog.rand,
                         rec.rand = rec.rand,
                         infection.FUN = infection.FUN,
                         recovery.FUN = recovery.FUN,
                         arrivals.FUN = arrivals.FUN,
                         departures.FUN = departures.FUN,
                         get_prev.FUN = get_prev.FUN)

  init <- init.icm(s.num = s.num,
                   e.num = e.num,
                   i.num = i.num,
                   q.num = q.num,
                   h.num = h.num,
                   r.num = r.num,
                   f.num = f.num)

  param <-  param.icm(inf.prob.e = inf.prob.e, 
                      act.rate.e = act.rate.e,
                      inf.prob.i = inf.prob.i, 
                      act.rate.i = act.rate.i,
                      inf.prob.q = inf.prob.q, 
                      act.rate.q = act.rate.q,                    
                      quar.rate = quar.rate,
                      hosp.rate = hosp.rate,
                      disch.rate = disch.rate,
                      prog.rate = prog.rate,
                      prog.dist.scale = prog.dist.scale,
                      prog.dist.shape = prog.dist.shape,
                      rec.rate = rec.rate,
                      rec.dist.scale = rec.dist.scale,
                      rec.dist.shape = rec.dist.shape,
                      fat.rate.base = fat.rate.base,
                      hosp.cap = hosp.cap,
                      fat.rate.overcap = fat.rate.overcap,
                      fat.tcoeff = fat.tcoeff,
                      vital = vital,
                      a.rate = a.rate, 
                      a.prop.e = a.prop.e,
                      a.prop.i = a.prop.i,
                      a.prop.q = a.prop.q,
                      ds.rate = ds.rate, 
                      de.rate = de.rate, 
                      di.rate = di.rate,
                      dq.rate = dq.rate,
                      dh.rate = dh.rate,
                      dr.rate = dr.rate)

  sim <- icm.seiqhrf(param, init, control)
  sim_df <- as.data.frame(sim, out=out)

  return(list(sim=sim, df=sim_df))
}
```

# Simulation runs

## Baseline

```{r, echo=TRUE, eval=TRUE}
baseline_sim <- simulate(ncores=4)
```

Visualise the results as a set of time-series of the daily count of our 100,000 individuals in each compartment.

```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
baseline_plot_df <- baseline_sim$df %>%
  # use only the prevalence columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  pivot_longer(-c(time),
               names_to="compartment",
               values_to="count") %>%
  filter(time <= 250)

# define a standard set of colours to represent compartments
compcols <- c("s.num" = "yellow", "e.num" = "orange", "i.num" = "red",
              "q.num" = "cyan", "h.num" = "magenta", "r.num" = "lightgreen",
              "f.num" = "black")
complabels <- c("s.num" = "Susceptible", "e.num" = "Infected/asymptomatic", 
                "i.num" = "Infected/infectious", "q.num" = "Isolated",
                "h.num" = "Requires hospitalisation", "r.num" = "Recovered",
                "f.num" = "Deaths due to COVID-19")

baseline_plot_df %>%
  # examine only the first 100 days since it
  # is all over by then using the default parameters
  filter(time <= 100) %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.7) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme_dark() +
    labs(title="Baseline simulation",
         x="Days since beginning of epidemic",
         y="Prevalence (persons)")
```

OK, that looks very reasonable. Note that almost the entire population ends up being infected. However, the **S** and **R** compartments dominate the plot (which is good, because it means humanity will survive!), so let's re-plot leaving out those compartments so we can see a bit more detail.

```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
baseline_plot_df %>%
  filter(compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  filter(time <= 100) %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.7) +
    scale_colour_manual(values = compcols, labels=complabels) +
    theme_dark() +
    labs(title="Baseline simulation",
         x="Days since beginning of epidemic",
         y="Prevalence (persons)")
```

Notice that the **I** compartment curve lags behind the **E** compartment curve -- the lag is the incubation period, and that the **Q** curve lags still further as infected people only reluctantly and belatedly self-isolate themselves (in this baseline scenario). 

## Running intervention experiments

### Scenario 1

Ramp up self-isolation rate, starting at day 1, ramp up to 0.3333, over 15 day period, then ongoing.

```{r, echo=TRUE, eval=TRUE}
isolation_ramp <- function(t) {
  ifelse(t <= 2, 0.0333, ifelse(t <= 17, 0.0333 + (t-3)*(0.3333 - 0.03333)/15, 0.3333))
}

exp1_isolation_ramp_sim <- simulate(quar.rate = isolation_ramp(1:366))  
```

### Scenario 2

Ramp up self-isolation rate, starting at day 1, ramp up to 0.6666, over 15 day period, then ongoing.

```{r, echo=TRUE, eval=TRUE}
big_isolation_ramp <- function(t) {
  ifelse(t <= 2, 0.0333, ifelse(t <= 17, 0.0333 + (t-3)*(0.6666 - 0.03333)/15, 0.6666))
}

exp2_big_isolation_ramp_sim <- simulate(quar.rate = big_isolation_ramp(1:366))  
```

### Scenario 3

As per scenario 1 plus a moderate increase in social distancing to 5 for everyone (halving the `act.rate`), again ramping it down between days 15 and 30, then maintain at 5 for a further 45 days, then revert to 10.

```{r, echo=TRUE, eval=TRUE}
social_distance_ramp <- function(t) {
  ifelse(t <= 17, 10, ifelse(t <= 32, 10 - (t-17)*(10 - 5)/15, ifelse(t <= 77, 5, 10)))
}

exp3_soc_dist_ramp_sim <- simulate(act.rate.i = social_distance_ramp(1:366),
                                   act.rate.e = social_distance_ramp(1:366),
                                   quar.rate = isolation_ramp(1:366))  
```

### Scenario 4

As per scenario 3 but four week lockdown starting at day 30.

```{r, echo=TRUE, eval=TRUE}
fourweek_lockdown_day30_vector <-c(rep(10, 32), rep(2.5, 30), rep(10, 304))

exp4_fourweek_lockdown_day30_sim <- simulate(act.rate.i = fourweek_lockdown_day30_vector,
                                             act.rate.e = fourweek_lockdown_day30_vector,
                                             quar.rate = isolation_ramp(1:366))  
```

### Scenario 5

As per scenario 4 but eight week lockdown starting at day 30.

```{r, echo=TRUE, eval=TRUE}
eightweek_lockdown_day30_vector <-c(rep(10, 32), rep(2.5, 60), rep(10, 274))

exp5_eightweek_lockdown_day30_sim <- simulate(act.rate.i = eightweek_lockdown_day30_vector,
                                             act.rate.e = eightweek_lockdown_day30_vector,
                                             quar.rate = isolation_ramp(1:366))  
```

### Scenario 6

As per scenario 4 but eight week lockdown starting at day 30, followed by self-isolation with high compliance (66% per day) on an ongoing basis.

```{r, echo=TRUE, eval=TRUE}
post_lockdown_isolation_ramp <- function(t) {
  ifelse(t <= 2, 0.0333, ifelse(t <= 17, 0.0333 + (t-3)*(0.3333 - 0.03333)/15, 
                                ifelse(t <= 92, 0.3333, 0.6666)))
}

exp6_eightweek_lockdown_day30_then_big_isol_sim <- simulate(act.rate.i = eightweek_lockdown_day30_vector,
                                             act.rate.e = eightweek_lockdown_day30_vector,
                                             quar.rate = post_lockdown_isolation_ramp(1:366))  
```

# Results

Now let's examine the results.

```{r, echo=TRUE, eval=TRUE, warning=FALSE, fig.height=12, fig.width=10}
elongate <- function(sim) {
  
  sim_df <- sim$df %>%
  # use only the prevalence columns
  select(time, s.num, e.num, i.num, q.num, 
         h.num, r.num, f.num) %>%
  # examine only the first 100 days since it
  # is all over by then using the default parameters
  filter(time <= 250) %>%
  pivot_longer(-c(time),
               names_to="compartment",
               values_to="count")
  return(sim_df)
}

exp1_isolation_ramp_sim_df <- elongate(exp1_isolation_ramp_sim)
exp2_big_isolation_ramp_sim_df <- elongate(exp2_big_isolation_ramp_sim)
exp3_soc_dist_ramp_sim_df <- elongate(exp3_soc_dist_ramp_sim)
exp4_fourweek_lockdown_day30_sim_df <- elongate(exp4_fourweek_lockdown_day30_sim )
exp5_eightweek_lockdown_day30_sim_df <- elongate(exp5_eightweek_lockdown_day30_sim)
exp6_eightweek_lockdown_day30_then_big_isol_sim_df <- elongate(exp6_eightweek_lockdown_day30_then_big_isol_sim)

combined_plot_df <- baseline_plot_df %>%
  mutate(experiment="Baseline") %>%
  bind_rows(exp1_isolation_ramp_sim_df %>%
              mutate(experiment="Scenario 1")) %>%
  bind_rows(exp2_big_isolation_ramp_sim_df %>%
              mutate(experiment="Scenario 2")) %>%
  bind_rows(exp3_soc_dist_ramp_sim_df %>%
              mutate(experiment="Scenario 3")) %>%
  bind_rows(exp4_fourweek_lockdown_day30_sim_df %>%
              mutate(experiment="Scenario 4")) %>%
  bind_rows(exp5_eightweek_lockdown_day30_sim_df %>%
              mutate(experiment="Scenario 5")) %>%
  bind_rows(exp6_eightweek_lockdown_day30_then_big_isol_sim_df %>%
              mutate(experiment="Scenario 6"))

p1 <- combined_plot_df %>%
  filter(compartment %in% c("e.num","i.num",
                            "q.num")) %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.7) +
    facet_grid(experiment ~ .) +
    scale_colour_manual(values = compcols, labels=complabels) +
    scale_y_continuous(labels = scales::comma) +
    theme_dark() +
    theme(legend.title = element_blank(),
          legend.position = "top",
          legend.direction = "horizontal") +
    labs(x="Days since beginning of epidemic",
         y="Prevalence (persons)")

p2 <- combined_plot_df %>%
  filter(compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.7) +
    geom_hline(yintercept=hc_scaler*40, colour="red") +
    annotate("text", x = 210, y = hc_scaler*(40 - 8.5), 
             label = "Hospital capacity", size=3) +
    facet_grid(experiment ~ .) +
    scale_colour_manual(values = compcols, labels=complabels) +
    scale_y_continuous(labels = scales::comma) +
    labs(x="Days since beginning of epidemic",
         y="") +
    theme_dark() +
    theme(legend.title = element_blank(),
          legend.position = "top",
          legend.direction = "horizontal")

p1 + p2
```

```{r, echo=FALSE, eval=TRUE, warning=FALSE}
toc()
```


